\textbf{ Spectrophotometric parallaxes with linear models:\\ Accurate distances for luminous red-giant stars }

### 2018-10-21 paper/sp.tex
rix comments

% To-do items
% -----------
% - need a snippet for Table 1.
% - check with WISE experts that we have the correct WISE photometry
% - ACE: What fraction of spectral pixels are (typically, or in each training set) zeroed out?

@@

\begin{abstract}\noindent
% context
With contemporary infrared spectroscopic surveys like \apogee,
red-giant stars can be observed to distances and extinctions
at which \gaia\ parallaxes are not highly informative.
Yet the combination of effective temperature, surface gravity, composition, and age---all
accessible through spectroscopy---determines a giant's luminosity.
Therefore spectroscopy plus photometry should enable 
precise spectrophotometric distance estimates.
% aims
Here we use the \apogee--\gaia--\zmass--\wise\ overlap to train a data-driven model
to predict parallaxes for red-giant branch stars with $0<\logg\leq2.2$.
% methods
We employ (the exponentiation of)
a linear function of \apogee\ spectral pixel intensities and multi-band photometry
to predict parallax spectrophotometrically.
The model training involves no logarithms or inverses of the \gaia\ parallaxes,
and needs no cut on the \gaia\ parallax signal-to-noise ratio.
It includes an L1 regularization to zero out the contributions of
uninformative pixels.
The training is performed
with leave-out subsamples such that no star's astrometry is used even indirectly in its
spectrophotometric parallax estimate.
% results
The model implicitly performs a reddening and extinction correction in its parallax prediction,
without any explicit dust model.
We assign to each star in the sample a new spectrophotometric parallax estimate;
these parallaxes have uncertainties of a few to 15 percent, depending on data quality,
which is more precise than the \gaia\ parallax for the vast majority of targets, and
certainly any stars more than a few kpc distance.
We obtain 10-percent distance estimates out to heliocentric distances of 20\,kpc,
and make global maps of the Milky Way's disk.
\end{abstract}

@@

As we operate these spectroscopic surveys,
the \gaia\ Mission (\citealt{gaia}) has also been revolutionizing our view of the disk.
It delivers good proper motions everywhere,
and extremely valuable parallax information locally.
However, it does not deliver precise geometric parallaxes over a large fraction of
the disk, and certainly not in the dusty and crowded regions.
For these reasons, the value of \gaia\ in these infrared spectroscopy projects is not
to deliver distance information directly, but rather to calibrate stellar models,
which then deliver distance information through relationships between stellar
luminosities and spectral characteristics.

Luminous red-giant stars are valuable for mapping the Milky Way disk for a number
of reasons, only one of which is their great luminosities (and hence brightnesses
even at large distance).
They are extremely common (frequent) stars, much more common than distance-indicating
stars like Cepheids.
They are produced by stellar populations at all metallicities and almost all ages,
so they can be found in all parts of the Galaxy.
They have temperatures and surface gravities such that it is
straightforward to spectroscopically measure metallicities,
detailed chemical abundances, and stellar parameters, even including ages (\citealt{martig, nessage}).
The red-giant branch is photometrically near-orthogonal to reddening vectors by dust,
so they are easy to de-redden or dust-correct.
And finally, in physical models of red giants, their predicted luminosities are simple
functions of composition, surface gravity, temperature, and age.
Thus, if we can measure these properties of red giants, we can (in principle) predict
their luminosities and use them as standard candles.
That is, red giants are standardizable candles.

In this project, we develop purely data-driven techniques to predict
red-giant parallaxes or distances with spectroscopy and photometry.
Our method capitalizes on the point that red-giant stars are dust-correctable,
standardizable candles, but it makes no use whatsoever of any stellar interior,
stellar exterior, or dust-reddening models.
We only use these physical expectations about red giants to recommend
data features---inputs---for a completely data-driven model.
The data-driven method uses patterns in the observed data themselves
to discover the relationships between spectral features in the spectra of the stars,
photometry (including colors), and parallax (or distance).

@@

Simple data-driven models (like the ones we will employ here)
also suffer from many limitations for this kind of work:
Because they are constrained only by the data, and not physical law,
they can learn relationships that we know
to be physically unlikely, and (by the same token) they can't benefit from our
physical knowledge.
They are required, in some sense, to spend some of the information in the data on
learning physical laws or relationships.
Not all of the information in the data is being used
optimally when we ignore our physical models.
Relatedly, they are not (usually) interpretable or generalizable:
A data-driven model trained with one kind of data can not usually be used in another
context with very different input data.
And the internal properties of the model are not useful, in general, for improving
our understanding or intuition about the astrophysics in play.

All that said, our goal here is to produce a precise mapping tool for surveys
like \apogee\ and \sdssv.
So we will take the limitations of the data-driven models in exchange for their
performance.
This work builds on the success we have seen in building other kinds of
data-driven tools for  stellar spectroscopy, for example to measure stellar
parameters (\citealt{cannon}), abundances (\citealt{ho, casey, nessdopp}),
and masses and ages (\citealt{nessage}).

There is a huge range of complexity or capacity in data-driven
models, from linear regressions (what we use here) to the most advanced machine-learning
methods (like deep neural networks).
In this work we stay on the low-complexity side of this.
We do not want to face the (literally combinatoric) choices in model structure that
advanced methods bring.
And we want to control or understand the smoothness or flexibility of the model.
Deep networks can model extremely complex function spaces, which is good in some
contexts, but bad when you know that the functional dependencies you hope to
model are smooth:
In that case, much of the information in the data can get spent on learning the
smoothness of the function.

What follows is a regression:
We use \gaia\ measurements of parallax to learn relationships between 
the spectroscopic and photometric properties of stars and their parallaxes.
We then use that learned model to estimate improved parallaxes for stars with
poor (or no) \gaia\ measurements.
This regression is very sensitive to certain aspects of the data or experimental
design.
One is that we will do far better with the regression if we have an accurate
model for noise process in the \gaia\ data.
For this reason, we make use of a justifiable likelihood function for the
\gaia\ astrometry (\citealt{gaialf}).
Another is that regressions can be strongly biased if we have a bad selection function
or data censoring.
This means that we cannot apply cuts to the \gaia\ astrometry or signal-to-noise
unless those cuts are correctly accounted for in our generative model for the
data sample.
Below, we will make no such cuts. That means that our training step has the odd
property that much of the training data is low in signal-to-noise and some of the
training parallax measurements are even negative!
The fact that we are using a justified likelihood function ensures that these
data will only affect the model in sensible and justifiable ways.
The alternative---cutting to high signal-to-noise data---is the more traditional
approach, but it leads to biased inferences (unless the cuts are included correctly
in the likelihood function).
The method in which we make no cuts and keep an umodified, justified likelihood function
is unbiased, and simple.

@@

In addition to making the above assumptions, we also avoid various practices
with the data that are tempting but lead to strong biases.
For example, we never cut on \gaia\ parallax or parallax signal-to-noise.
The common practices of cutting to parallaxes that are good to 20~percent,
or Catalog parallaxes that are positive, or parallaxes that are smaller or
larger than something, are all practices that will bias results on stellar
collections.
That is, if you cut on parallax or parallax signal-to-noise and you subsequently
take an average of parallaxes for some population, or perform a regression (as we
do here), the results of that average or regression will be strongly biased.
We never cut on parallax or parallax signal-to-noise.
By using a justifiable likelihood function (the ``likelihood'' assumption above),
we can use all of the \gaia\ parallaxes without the low signal-to-noise and
negative parallaxes causing trouble for our regressions, and without the biases
that enter when cuts are made.

@@

The model and log-likelihood function can be expressed heuristically as
\begin{eqnarray}
\gparallax_n &=& \exp(\theta\cdot x_n) + \mbox{noise}
\\
\chi^2(\theta) &\equiv& \sum_{n=1}^N \frac{[\gparallax_n - \exp(\theta\cdot x_n)]^2}{\gsigma_n^2}
\end{eqnarray}
where
$\gparallax_n$ is the \gaia\ measurement (adjusted; see below) of the astrometric parallax of star $n$,
the model is that the logarithm of the true parallax
can be expressed as a linear combination of the components
of some $D$-dimensional feature vector $x_n$,
$\theta$ is a $D$-vector of linear coefficients,
$\gsigma_n$ is the \gaia\ estimate of the uncertainty on the parallax measurement,
and $\chi^2(\theta)$ is (twice) the negative-log-likelihood for the parameters $\theta$
under the assumption of known Gaussian noise and
that there are $N$ independently measured stars $n$.
The feature vector $x_n$ contains photometry in a few bands, and all 7400-ish pixels
in the pseudo-continuum-normalized \apogee\ spectrum, so $D$ is on the order of 7400.

@@

This optimization is not convex---there are multiple optima in general.
In particular, there is a large, degenerate, pathological optimum where
the exponential function underflows, all parallaxes are predicted to vanish,
and there is no gradient of anything with respect to the parameter $D$-vector $\theta$.
This large, bad optimum (it is a local, not global, optimum) must be avoided in optimization.
In practice we avoid it by optimizing first for the very highest signal-to-noise
(in astrometric parallax) Training Set stars, and then use that first optimum as a
good first guess or initialization for the full optimization.
This method works because in the limit of high signal-to-noise, the problem asymptotically approaches
the convex problem of L1-regularized linear least-square fitting.
Optimization is performed with \code{scipy.optimize} using the
\acronym{L-BFGS-B} algorithm (\citealt{lbfgsb}).

Once the model is optimized, the output of the model is a prediction of the parallax,
or really what we will call the \emph{spectrophotometric parallax}.
This spectrophotometric parallax
$\sparallax_m$ for any star $m$ in the validation or test set is
assigned according to
\begin{eqnarray}
\sparallax_m &\leftarrow& \exp(\hat{\theta}\cdot x_m)
\end{eqnarray}
where
$\hat{\theta}$ is the optimal parameter vector according
to \equationname~(\ref{eq:objective}),
and
$x_m$ is the feature $D$-vector for star $m$.
The model can be trained on a training set of stars and applied to
a validation set or a test set to make predictions.
The only requirements are that every test-set object $m$  must have a full feature
vector $x_m$ just as every training-set object $n$ must have a full feature
vector $x_n$.

@@

One final detail: We perform all of the fitting in a two-fold train-and-test framework,
in which the data are split into two disjoint subsets, A and B.
The model trained on the A data is used to predict or produce
the spectrophotometric parallaxes $\sparallax_m$ values (and hence
the distances and parallaxes) of the B data,
and the model trained on the B data is used to 
produce the $\sparallax_m$ values of the A data.
This ensures that, in the estimate of any individal star's
spectrophotometric parallax, none of the \gaia\ data pertaining to that star were used.
This makes the parallax estimates from the spectrophotometric feature vectors
$x_m$ statistically independent of the \gaia\ parallax measurements.
They are not independent globally---\gaia\ data were used to train the model---but
each star's spectrophotometric parallax estimate is independent
of the astrometric estimate from \gaia\ on a star-by-star basis.

@@

Every Parent-Sample star gets, in addition, a randomly assigned binary
label (A or B).
This is used for two-fold validation and jackknife.
In short, we will
use the model trained on the Training Set A data to assign spectrophotometric parallaxes
to the Parent Sample B data and
use the model trained on the Training Set B data to assign spectrophotometric parallaxes
to the Parent Sample A data.

For training, we need A and B Training Sets.
We define the Training Set stars to be Parent-Sample stars that also
have a measured \gaia\ parallax.
The Training-Set stars, in addition to meeting all Parent-Sample cuts,
also must meet an uncertainty criterion $\gsigma_n < 0.1\,\mathrm{mas}$, i.e. 
\begin{equation}
\code{parallax_error < 0.1}
\end{equation}
it must be observed by more than $8$ (widely separated) times
\begin{equation}
\code{visibility_periods_used >= 8}
\end{equation}
as well as meet the additional quality criterion
\begin{equation}
\code{astrometric_chi2_al / sqrt(astrometric_n_good_obs_al - 5) <= 35}
\end{equation}
which eliminates detections that are not consistent with a single PSF \citep[see][]{calj},
to ensure that the parallax measurements are good.
These cuts leave 28226 stars total in the Training Sets.
But---as we have emphasized above---we do not cut ever on parallax or
or the ratio to its uncertainty. For this reason, most of the Training-Set stars
do not have significantly measured parallaxes and more than 2~percent even have
negative parallaxes!
But we will perform our training such that it will be unbiased in these
circumstances.
Indeed it would be necessarily biased if ever we did cut on parallax or
parallax signal-to-noise.

In detail,
for each star $n$ in the Training Sets, we take from \gaia\ \acronym{DR2} (\citealt{gaiadr2})
the astrometric parallax $\gparallax_n$ and its uncertainty $\gsigma_n$.
Importantly, to each \gaia\ parallax measurement we add a positive
offset $\varpi_0$ to adjust for the under-estimation of
parallaxes reported by the \gaia\ Collaboration (\citealt{lindegren}).
The \gaia\ recommended value is $\varpi_0=0.029$\,mas (\citealt{lindegren}),
but we adopt $\varpi_0 = 0.048$\,mas because that value
optimizes our cross-validation objective.
It also happens to be in the range of other values found in the literature (\citealt{arenou, zinn}).
Of course we really expect the offset to be a function of sky position and color (at least),
so the offset we find is not recommended for further use as any kind of universal value.
The spectrophotometric parallaxes we generate
do not depend strongly on this choice, except
for small differences at very small parallax (large distance).
The Training Sets are shown in the bottom panels of \figurename~\ref{fig:samples}.

@@

We optimize two models, one for Training Set A and one for Training Set B.
The two models are applied to the two splits of the Parent Sample, Parent Sample
A and Parent Sample B, using the A-trained model on Parent Sample B and
\foreign{vice versa}.
For the purposes of assessing the accuracy and precision of the model, this
train-and-test framework constitutes a two-fold cross-validation.
Results of this cross-validation is shown in \figurename~\ref{fig:xval}.
For the stars with highest \gaia-measured parallax signal-to-noise,
the spectrophotometric model is predicting the \gaia\ parallaxes with little bias
and a scatter of less than 9~percent.
\begin{figure}
\includegraphics[width=\textwidth]{residuals.pdf}
\caption{The predictive accuracy of the model, in the two-fold cross-validation.
  Each panel shows \gaia\ astrometric parallaxes $\gparallax_n$ on the vertical axis
  and our spectrophotometric parallaxes $\sparallax$ on the horizontal axis.
  The left panel shows the full Parent Sample, the middle panel shows the Training Set
  stars (which pass \gaia\ quality checks; see text), and the right panel shows a
  subset of stars with very high astrometric parallax signal-to-noise. This latter
  set plays no particular role in the method, but it can be used to demonstrate or
  assess the prediction precision. The fractional precision of the prediction in the
  right panel is better than 9 percent. Note that by construction, there can be no negative
  spectrophotometric parallaxes; that is, the quantity plotted on the vertical axis is
  positive-definite.\label{fig:xval}}
\end{figure}

@@

This 9-percent precision estimate is conservative in the sense that it does not
deconvolve or correct for the contributions from the \gaia\ noise.
However it is also not fully conservative, in that the test is performed with
\gaia's best stars, which are bright, nearby, in low extinction regions, and
(because of these things) near Solar metallicity.
There could in principle be additional bias or scatter for stars that are in dustier
regions or at lower metallicities.
\figurename~\ref{fig:residuals} shows that
there is no suggestion of any such trends when we color the residuals by metallicity
or $H-W_2$ color (which is a reddening proxy).
However, that figure does show that stars with greater reddening do appear to have
larger scatter against the \gaia\ parallaxes.
Additional tests confirm that our spectrophotometric parallaxes have largest scatter
in the dustiest and most crowded regions, as expected given the relatively low angular resolution
photometry.
\begin{figure}
\includegraphics[width=\textwidth]{./residuals_training.pdf}
\caption{Prediction validation colored by relevant features.
  Each panel shows the predicted parallax \foreign{vs} the
  spectrophotometric parallax colored by a different
  data feature that might be relevant to the residual.
  Again the quantity plotted on vertical axis is positive-definite.\label{fig:residuals}}
\end{figure}

Another validation of the results can be obtained by looking at known
clusters or spatially compact objects in the data.
In \figurename~\ref{fig:clusters}, we show the parallax distribution from \gaia\ and
the spectrophotometric-parallax distribution from our model for stars that are
astrometrically confirmed members of three of the stellar clusters in which we
validated our results.
The cluster members were found by hand, looking at sky and proper-motion distributions
centered on literature values.
In each case we are plotting only very securely identified cluster members; that
is, membership is conservative.
We chose three clusters to span some range in metallicity and age, and therefore test
the accuracy of the method for different kinds of populations.
With the exception of \acronym{M71}, the clusters we tested
indicate that both \gaia\ and the spectrophotometric parallaxes appear to be unbiased
for these clusters and the range in abundances and ages they represent.
The case of cluster \acronym{M71} is troubling, but after searching for trends with
housekeeping data (also see \figurename~\ref{fig:residuals}) we don't find any correlations
with parallax offsets.
\begin{figure}
~\hfill\includegraphics[width=0.7\textwidth]{clusters.pdf}\hfill~
\caption{The \gaia\ astrometric parallaxes $\gparallax_n$
  and spectrophotometric parallaxes $\sparallax_n$ for member stars for
  three of the stellar clusters in the \apogee\ observing footprint. The spectrophotometric
  parallaxes and their inverse-variance-weighted mean and the astrometric parallaxes
  and their inverse-variance-weighted mean
  are shown as vertical lines.
  Cluster members were identified to be close to the cluster in celestial coordinates
  and comoving in proper motion.\label{fig:clusters}}
\end{figure}

@@

This project demonstrates that photometry and spectroscopy can be used to predict
or estimate distances to stars.
This isn't new!
However, what's new is that it is possible to do this very well but with a purely
linear model acting on magnitudes and spectral (logarithmic) fluxes, and entirely
data-driven.
That is, a linear model trained on noisy parallax data from \gaia.

@@

However, our choice of features is very rigid:
In principle the data themselves should tell us what features to include.
In that direction, it would make sense to choose features using a more complex
technique like deep learning or an auto-encoder.
These methods generate good features automatically.
However, they also involve an enormous set of hyper-parameter-like choices,
and they involve sacrificing certain kinds of interpretability.
However, we expect that a better set of features would do better; in this sense
the project here is just a first step towards precise spectrophotometric distance
estimates.

>>>>

% To-do items
% -----------
% - check with WISE experts that we have the correct WISE photometry
% - ACE: What fraction of spectral pixels are (typically, or in each training set) zeroed out?
% - are we clear that the map is just a demo and just shows the 10-percent-ish region of the MW disk?

@@

\begin{abstract}\noindent
% context
With contemporary infrared spectroscopic surveys like \apogee,
red-giant stars can be observed to distances and extinctions
at which \gaia\ parallaxes are not highly informative.
Yet the combination of effective temperature, surface gravity, composition, and age---all
accessible through spectroscopy---determines a giant's luminosity.
Therefore spectroscopy plus photometry should enable 
precise spectrophotometric distance estimates.
% aims
Here we use the \apogee--\gaia--\zmass--\wise\ overlap to train a data-driven model
to predict parallaxes for red-giant branch stars with $0<\logg\leq2.2$ (more luminous
than the red clump).
% methods
We employ (the exponentiation of)
a linear function of \apogee\ spectral pixel intensities and multi-band photometry
to predict parallax spectrophotometrically.
The model training involves no logarithms or inverses of the \gaia\ parallaxes,
and needs no cut on the \gaia\ parallax signal-to-noise ratio.
It includes an L1 regularization to zero out the contributions of
uninformative pixels.
The training is performed
with leave-out subsamples such that no star's astrometry is used even indirectly in its
spectrophotometric parallax estimate.
% results
The model implicitly performs a reddening and extinction correction in its parallax prediction,
without any explicit dust model.
We assign to each star in the sample a new spectrophotometric parallax estimate;
these parallaxes have uncertainties of a few to 15 percent, depending on data quality,
which is more precise than the \gaia\ parallax for the vast majority of targets, and
certainly any stars more than a few kpc distance.
We obtain 10-percent distance estimates out to heliocentric distances of 20\,kpc,
and make global maps of the Milky Way's disk.
\end{abstract}

@@

As we operate these spectroscopic surveys,
the \gaia\ Mission (\citealt{gaia}) has also been revolutionizing our view of the disk.
It delivers good proper motions over most of the Galaxy,
and extremely valuable parallax information locally.
However, \gaia\ does not deliver precise geometric parallaxes over a large fraction of
the disk, and certainly not in the dusty and crowded regions.
For these reasons, the value of \gaia\ in these infrared spectroscopy projects is not
to deliver distance information directly, but rather to calibrate stellar models,
which then deliver distance information through relationships between stellar
luminosities and spectral characteristics.

Luminous red-giant stars are valuable for mapping the Milky Way disk for a number
of reasons, only one of which is their great luminosities (and hence brightnesses
even at large distance).
They are very common stars, much more frequent than distance-indicating
stars like Cepheids.
They are produced by stellar populations at all metallicities and almost all ages,
so they can be found in all parts of the Galaxy.
They have temperatures and surface gravities such that it is
straightforward to spectroscopically measure metallicities,
detailed chemical abundances, and stellar parameters, even including ages (\citealt{martig, nessage}).
The red-giant branch is photometrically near-orthogonal to reddening vectors by dust,
so they are easy to de-redden or dust-correct.
And finally, in physical models of red giants, their predicted luminosities are simple
functions of composition, surface gravity, temperature, and age.
Thus, if we can measure these properties of red giants, we can (in principle) predict
their luminosities and use them as standard candles.
That is, red giants are standardizable candles.

In this project, we develop purely data-driven techniques to predict
red-giant parallaxes or distances with spectroscopy and photometry.
Our method is predicated on the expectation that red-giant stars are dust-correctable,
standardizable candles with such data,
but it makes no use whatsoever of any stellar interior,
stellar exterior, or dust-reddening models.
We only use these physical expectations about red giants to recommend
data features---inputs---for a completely data-driven model.
The data-driven method uses patterns in the observed data themselves
to discover the relationships between spectral features in the spectra of the stars,
photometry (including colors), and parallax (or distance).

@@

Simple data-driven models (like the ones we will employ here)
also suffer from many limitations for this kind of work:
Because they are constrained only by the data, and not physical law,
they can learn relationships that we know
to be physically unlikely, and (by the same token) they can't benefit from our
physical knowledge.
They are required, in some sense, to spend some of the information in the data on
learning physical laws or relationships.
Not all of the information in the data is being used
optimally when we ignore our physical models.
Relatedly, they are not (usually) interpretable or generalizable or useful for
extrapolations significantly outside the training set:
A data-driven model trained with one kind of data can not usually be used in another
context with very different input data.
And the internal properties of the model are not useful, in general, for improving
our understanding or intuition about the astrophysics in play.

All that said, our goal here is to produce a precise mapping tool for surveys
like \apogee\ and \sdssv.
So we accept the limitations of the data-driven models in exchange for their
performance.
This work builds on the success we have seen in building other kinds of
data-driven tools for  stellar spectroscopy, for example to measure stellar
parameters (\citealt{cannon}), abundances (\citealt{ho, casey, nessdopp}),
and masses and ages (\citealt{nessage}).

There is a huge range of complexity or capacity in data-driven
models, from linear regressions (what we use here) to the most advanced machine-learning
methods (like deep neural networks).
In this work we stay on the low-complexity side of this, as
we do not want to face the (literally combinatoric) choices in model structure that
advanced methods bring.
And we want to control or understand the smoothness or flexibility of the model.
Deep networks can model extremely complex function spaces, which is good in some
contexts, but bad when you know that the functional dependencies you hope to
model are smooth:
In that case, much of the information in the data can get spent on learning the
smoothness of the function.

What follows is a regression:
We use \gaia\ measurements of parallax to learn relationships between 
the spectroscopic and photometric properties of stars and their parallaxes.
We then use that learned model to estimate improved parallaxes for stars with
poor (or no) \gaia\ measurements.
This regression is very sensitive to certain aspects of the data or experimental
design.
One is that we will do far better with the regression if we have an accurate
model for noise process in the \gaia\ data.
For this reason, we make use of a justifiable likelihood function for the
\gaia\ astrometry (\citealt{gaialf}).
Another is that regressions can be strongly biased if we have a bad selection function
or data censoring.
This means that we cannot apply cuts to the \gaia\ astrometry or signal-to-noise
unless those cuts are correctly accounted for in our generative model for the
data sample.
Below, we will make no such cuts. That means that our training step has the odd
property that much of the training data is low in signal-to-noise; we include
training parallax measurements even if they are negative!
The fact that we are using a justified likelihood function ensures that these
data will only affect the model in sensible and justifiable ways.
The alternative---cutting to high signal-to-noise data---is the more traditional
approach, but it leads to biased inferences (unless the cuts are included correctly
in the likelihood function).
The method in which we make no cuts and keep an umodified, justified likelihood function
is unbiased, and simple.

@@

In addition to making the above assumptions, we also avoid various practices
with the data that are tempting but lead to important biases.
For example, we never cut on \gaia\ parallax or parallax signal-to-noise.
The common practices of cutting to parallaxes that are good to 20~percent
(see, for example, \citealt{trick, helmi}),
or Catalog parallaxes that are positive, or parallaxes that are smaller or
larger than something, are all practices that will bias results on stellar
collections.
That is, if you cut on parallax or parallax signal-to-noise and you subsequently
take an average of parallaxes for some population, or perform a regression (as we
do here), the results of that average or regression will be strongly biased.
We never cut on parallax or parallax signal-to-noise.
By using a justifiable likelihood function (the ``likelihood'' assumption above),
we can use all of the \gaia\ parallaxes without the low signal-to-noise and
negative parallaxes causing trouble for our regressions, and without the biases
that enter when cuts are made.

@@

The model for the parallax
and the log-likelihood function can be expressed heuristically as
\begin{eqnarray}
\gparallax_n &=& \exp(\theta\cdot x_n) + \mbox{noise}
\\
\chi^2(\theta) &\equiv& \sum_{n=1}^N \frac{[\gparallax_n - \exp(\theta\cdot x_n)]^2}{\gsigma_n^2}
\end{eqnarray}
where
$\gparallax_n$ is the \gaia\ measurement (adjusted; see below) of the astrometric parallax of star $n$,
the model is that the logarithm of the true parallax
can be expressed as a linear combination of the components
of some $D$-dimensional feature vector $x_n$,
$\theta$ is a $D$-vector of linear coefficients,
$\gsigma_n$ is the \gaia\ estimate of the uncertainty on the parallax measurement,
and $\chi^2(\theta)$ is (twice) the negative-log-likelihood for the parameters $\theta$
under the assumption of known Gaussian noise and
that there are $N$ independently measured stars $n$.
The feature vector $x_n$ contains photometry in a few bands, and all 7400-ish pixels
in the pseudo-continuum-normalized \apogee\ spectrum, so $D$ is on the order of 7400.

@@

This optimization is not convex---there are multiple optima in general.
In particular, there is a large, degenerate, pathological optimum where
the exponential function underflows, all parallaxes are predicted to vanish,
and there is no gradient of anything with respect to the parameter $D$-vector $\theta$.
This large, bad optimum (it is a local, not global, optimum) must be avoided in optimization.
In practice we avoid it by optimizing first for the very highest signal-to-noise
(better than 5-percent measurements of astrometric parallax)
Training Set stars, and then use that first optimum as a
good first guess or initialization for the full optimization.
This method works because in the limit of high signal-to-noise, the problem asymptotically approaches
the convex problem of L1-regularized linear least-square fitting.
Optimization is performed with \code{scipy.optimize} using the
\acronym{L-BFGS-B} algorithm (\citealt{lbfgsb}).

Once the model is optimized---and therefore $\hat{\theta}$ is determined---the
output of the model is a prediction of the parallax,
or really what we will call the \emph{spectrophotometric parallax}.
This spectrophotometric parallax
$\sparallax_m$ for any star $m$ in the validation or test set is
assigned according to
\begin{eqnarray}
\sparallax_m &\leftarrow& \exp(\hat{\theta}\cdot x_m)
\end{eqnarray}
where
$\hat{\theta}$ is the optimal parameter vector according
to \equationname~(\ref{eq:objective}),
and
$x_m$ is the feature $D$-vector for star $m$.
The model can be trained on a training set of stars and applied to
a validation set or a test set to make predictions.
The only requirements are that every test-set object $m$  must have a full feature
vector $x_m$ just as every training-set object $n$ must have a full feature
vector $x_n$.

@@

We perform all of the fitting in a two-fold train-and-test framework,
in which the data are split into two disjoint subsets, A and B.
The model trained on the A data is used to predict or produce
the spectrophotometric parallaxes $\sparallax_m$ values (and hence
the distances and parallaxes) of the B data,
and the model trained on the B data is used to 
produce the $\sparallax_m$ values of the A data.
This ensures that, in the estimate of any individal star's
spectrophotometric parallax, none of the \gaia\ data pertaining to that star were used.
This makes the parallax estimates from the spectrophotometric feature vectors
$x_m$ statistically independent of the \gaia\ parallax measurements.
They are not independent globally---\gaia\ data were used to train the model---but
each star's spectrophotometric parallax estimate is independent
of the astrometric estimate from \gaia\ on a star-by-star basis.

@@

Every Parent Sample star gets, in addition, a randomly assigned binary
label (A or B).
This is used for two-fold validation and jackknife.
In short, we will
use the model trained on the Training Set A data to assign spectrophotometric parallaxes
to the Parent Sample B data and
use the model trained on the Training Set B data to assign spectrophotometric parallaxes
to the Parent Sample A data.

For training, we need A and B Training Sets.
We define the Training Set stars to be Parent Sample stars that also
have a measured \gaia\ parallax.
The Training Set stars, in addition to meeting all Parent Sample cuts,
also must meet an uncertainty criterion $\gsigma_n < 0.1\,\mathrm{mas}$, i.e. 
\begin{equation}
\code{parallax_error < 0.1}
\end{equation}
it must be observed by more than $8$ (widely separated) times
\begin{equation}
\code{visibility_periods_used >= 8}
\end{equation}
as well as meet the additional quality criterion
\begin{equation}
\code{astrometric_chi2_al / sqrt(astrometric_n_good_obs_al - 5) <= 35}
\end{equation}
which eliminates detections that are not consistent with a single PSF \citep[see][]{calj},
to ensure that the parallax measurements are good.
These cuts leave 28226 stars total in the Training Sets.
But---as we have emphasized above---we do not cut ever on parallax or
or the ratio to its uncertainty. For this reason, most of the Training Set stars
do not have significantly measured parallaxes and more than 2~percent even have
negative parallaxes!
But we will perform our training such that it will be unbiased in these
circumstances.
Indeed it would be necessarily biased if ever we did cut on parallax or
parallax signal-to-noise.

In detail,
for each star $n$ in the Training Sets, we take from \gaia\ \acronym{DR2} (\citealt{gaiadr2})
the astrometric parallax $\gparallax_n$ and its uncertainty $\gsigma_n$.
Importantly, to each \gaia\ parallax measurement we add a positive
offset $\varpi_0$ to adjust for the under-estimation of
parallaxes reported by the \gaia\ Collaboration (\citealt{lindegren}).
The \gaia\ recommended all-sky-average value is $\varpi_0=0.029$\,mas (\citealt{lindegren}),
but we adopt $\varpi_0 = 0.048$\,mas because that value
optimizes our cross-validation objective.
It also happens to be in the range of other values found in the literature (\citealt{arenou, zinn}).
Of course we really expect the offset to be a function of sky position and color (at least),
so the offset we find is not recommended for further use as any kind of universal value.
The spectrophotometric parallaxes we generate
do not depend strongly on this choice, except
for small differences at very small parallax (large distance).
The Training Sets are shown in the bottom panels of \figurename~\ref{fig:samples}.

@@

We optimize two models, one for Training Set A and one for Training Set B.
The two models are applied to the two splits of the Parent Sample, Parent Sample
A and Parent Sample B, using the A-trained model on Parent Sample B and
\foreign{vice versa}.
For the purposes of assessing the accuracy and precision of the model, this
train-and-test framework constitutes a two-fold cross-validation.
Results of this cross-validation are shown in \figurename~\ref{fig:xval}.
For the stars with highest \gaia-measured parallax signal-to-noise,
the spectrophotometric model is predicting the \gaia\ parallaxes with little bias
and a scatter of less than 9~percent.
\begin{figure}
\includegraphics[width=\textwidth]{residuals.pdf}
\caption{The predictive accuracy of the model, in the two-fold cross-validation.
  Each panel shows \gaia\ astrometric parallaxes $\gparallax_n$ on the vertical axis
  and our spectrophotometric parallaxes $\sparallax$ on the horizontal axis.
  The left panel shows the full Training Set (both A and B combined),
  and the right panel shows a
  subset of stars with very high astrometric parallax signal-to-noise. This latter
  set plays no particular role in the method, but it can be used to demonstrate or
  assess the prediction precision. The fractional precision of the prediction in the
  right panel is better than 9 percent. Note that by construction, there can be no negative
  spectrophotometric parallaxes; that is, the quantity plotted on the vertical axis is
  positive-definite.\label{fig:xval}}
\end{figure}

@@

This 9-percent precision estimate is conservative in the sense that it does not
deconvolve or correct for the contributions from the \gaia\ noise.
However, this test is performed with
\gaia's best stars, which are bright, nearby, in low extinction regions, and
(because of these things) near Solar metallicity.
There could in principle be additional bias or scatter for stars that are in dustier
regions or at lower metallicities.
\figurename~\ref{fig:residuals} shows that
there is no suggestion of any such trends when we color the residuals by metallicity
or $H-W_2$ color (which is a reddening proxy).
However, that figure does show that stars with greater reddening do appear to have
larger scatter against the \gaia\ parallaxes.
Additional tests confirm that our spectrophotometric parallaxes have largest scatter
in the dustiest and most crowded regions, as expected given the relatively low angular resolution
photometry.
\begin{figure}
\includegraphics[width=\textwidth]{./residuals_training.pdf}
\caption{Prediction validation colored by relevant features.
  Each panel shows the predicted parallax \foreign{vs} the
  spectrophotometric parallax colored by a different
  data feature that might be relevant to the residual.
  Again the quantity plotted on vertical axis is positive-definite.\label{fig:residuals}}
\end{figure}

Another validation of the results can be obtained by looking at known
clusters or spatially compact objects in the data.
In \figurename~\ref{fig:clusters}, we show the parallax distribution from \gaia\ and
the spectrophotometric-parallax distribution from our model for stars that are
astrometrically confirmed members of three of the stellar clusters in which we
validated our results.
The cluster members were found by hand, looking at sky and proper-motion distributions
centered on literature values.
In each case we are plotting only very securely identified cluster members; that
is, membership is conservative.
We chose the three displayed clusters
to span some range in metallicity and age and extinction, and therefore test
the accuracy of the method for different kinds of populations.
With the exception of \acronym{M71}, the clusters we tested
indicate that both \gaia\ and the spectrophotometric parallaxes appear to be unbiased
for these clusters and the range in abundances and ages they represent.
The case of cluster \acronym{M71} is troubling, but after searching for trends with
housekeeping data (also see \figurename~\ref{fig:residuals}) we don't find any correlations
with parallax offsets, although \acronym{M71} is a higher-extinction cluster.
\begin{figure}
~\hfill\includegraphics[width=0.7\textwidth]{clusters.pdf}\hfill~
\caption{The \gaia\ astrometric parallaxes $\gparallax_n$
  and spectrophotometric parallaxes $\sparallax_n$ for member stars for
  three of the stellar clusters in the \apogee\ observing footprint. The spectrophotometric
  parallaxes and their inverse-variance-weighted mean and the astrometric parallaxes
  and their inverse-variance-weighted mean
  are shown as vertical lines.
  Cluster members were identified to be close to the cluster in celestial coordinates
  and comoving in proper motion.\label{fig:clusters}}
\end{figure}

@@

We have demonstrated that photometry and spectroscopy can be used to predict
or estimate distances to stars.
This isn't new!
However, what's new is that it is possible to do this very well for luminous giants
with unknown ages, and with a purely
linear model acting on magnitudes and spectral (logarithmic) fluxes, and entirely
data-driven.
That is, a linear model trained on noisy parallax data from \gaia.

@@

However, our choice of features is very rigid:
In principle the data themselves should tell us what features to include.
In that direction, it would make sense to choose features using a more complex
technique like deep learning or an auto-encoder.
These methods generate good features automatically.
They also involve an enormous set of hyper-parameter-like choices,
and they involve sacrificing certain kinds of interpretability.
All that said, we expect that a better set of features would do better; in this sense
the project here is just a first step towards precise spectrophotometric distance
estimates.
